Can dike grassland restoration profit from ideal seed mixture-substrate combinations?

### 2023-02-17 manuscript.Rmd
Update manuscript.Rmd

```{r set-up, include = FALSE, message = FALSE}
library(here)
library(knitr)
library(officedown)
library(officer)
library(tidyverse)
library(flextable)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  dpi = 300,
  fig.cap = FALSE
  )
set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 11,
  text.align = "center"
  )
```

@@

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Though, adapted seed mixtures to certain site conditions promise a more effective restoration. This is especially important if the restored vegetation has to protect constructions like dikes. An ideal seed mixture-substrate combination would concurrently improve the ecosystem function erosion protection and enhance species-rich grasslands by a fast and successful establishment of target species.
2.  An experiment with 288 plots was conducted on both expositions of a dike and was surveyed 2018--2021. We tested different sand admixtures, substrate depths, seed densities and seed mixture types (hay meadows, dry calcareous grasslands), and measured the persistence of sown species, the Favorable Conservation Status (FCS) and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a significant but slightly positive effect on all indices. Substrate depth, seed density and seed mixture type had no significant effect on persistence or FCS.
4.  On the north exposition, hay meadow-seed mixtures developed closer to its reference than dry grasslands. Overall, the vegetation developed in the desired direction but a recovery debt remained after four years. Plots still exhibited similarities to negative references from ruderal vegetation.
5.  The south exposition was significantly worse than the north exposition which might be due to the severe drought during establishment and the addition and removal of a geotextile. This initial negative effect remained over the whole observation period.
6.  Synthesis and applications: Adapted seed mixtures can make restorations more efficient like applying hay meadows on the north exposition. However, varying seed mixture-substrate combinations might foster biodiversity like beta diversity more than a desired 'ideal' combination. Furthermore, additional efforts in management might be necessary to reduce the recovery debt and resowing after a severe drought might be sensible due to strong legacy effects.

@@

@@

Grasslands provide several ecosystem services and a high biodiversity [@Dengler.2014; @Bardgett.2021]. However, they are globally endangered [@Bardgett.2021]even by restoration through unsuited afforestation [@dudley2020]. In Europe, calcareous grasslands and hay meadows are on the red list [Category 3, vulnerable, @Janssen.2016]. Restoration is seen as a key factor to sustain ecosystem services and biodiversity [@COP.2014; @UnitedNations.2019b] and sowing is a long established restoration approach for species-rich grasslands [@Kiehl.2010]. Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts [@Freitag.2021]. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021].

@@

Dikes are promising sites for the restoration of species-rich grasslands, because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in an agricultural landscape [@Batori.2020]. Dikes have specific characteristics with strongly exposed slopes and river and land orientation [@Batori.2016] and need a vegetation cover to ensure dike stability [@berendse2015]. Dikes can reconcile several requirements [@teixeira2022], which can be fostered by an adapted seed mixture-substrate combination.

@@

The seed mixture-substrate combination were tested with an experiment on a dike of the River Danube in SE Germany, which was established in March 2018 (Figure \@ref(fig:map)); 314 m above sea level; WGS84 (lat/lon), 48.83895, 12.88412). The climate is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], and three minor floods which did not reach the plots [Appendix A1, @lfu2021waterlevel; @lfu2021discharge]. Regional sand (0--4 mm) was used to lean the substrate and the soil was taken from removed agricultural soil of a dike construction site near Steinkirchen. The substrates were mixed with a big roller and an excavator prepared the plots. The target vegetation were lowland hay meadows [EUNIS code: R22, @Chytry.2020; Arrhenatherion elatioris, EuroVegChecklist: CM01A, @mucina2016] and semi-dry calcareous grassland (R1A; Cirsio-Brachypodion pinnati, DA01B). The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds. From the species pools, 20 species were randomly selected for each plot (Appendix A2). The seed mixtures always contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Figure \@ref(fig:traits)). As expected, the hay meadows mixtures have a higher specific leaf area (SLA) lower seed mass and higher canopy height than the dry meadows mixtures. The south exposed plots were sown in mid-April 2018 and the north exposition two weeks later. Due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks.

The management started with a cut without hay removal and a mowing height of 20 cm in August 2018, followed by 'normal' deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions.

We used 288 plots of the size 2.0 m × 2.0 m which were distributed over the north and south exposition, halfway up the slopes of the dike and arranged in six blocks (= replicates). The experiment used a split plot design combined with randomized complete block design (Figure \@ref(fig:map)). The split plot was caused by testing all 24 treatments on both sides of the dike. We tested three sand admixtures (0%, 25% and 50%) and tow soil depths (15 vs. 30 cm) of practical interest [@KleberLerchbaumer.2017] and two . Below the substrate, a drainage layer of 5 cm consisting of gravel (0--16 mm) was installed. We tested soil samples of the three substrates on both expositions by mixing several sub-samples. The sand admixture changed the soil texture, increased the C/N ratio, but reduced Calciumcarbonat, and hardly changed the rather high pH values (Table \@ref(tab:substrate)). Phosphate and potassium were rather scarce for agricultural soils and Magnesium rather high [@lfl2022]. The two seed mixture types were tested with the normal seeding density and the increased density recommended for slopes [4 vs. 8 g m^-2^ @Kiehl.2010; @KleberLerchbaumer.2017]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. Establishment rates of species (Appendix A3).

@@

Reference vegetation surveys were extracted from sPlotOpen [@Sabatini.2021] and our own surveys in the surroundings on Danube dikes [@bauer2023]. We selected six calcareous grassland plots [EUNIS code R1A, @Chytry.2020] within SE Germany from sPlotOpen and 98 plots of our own survey which included furthermore lowland hay meadows (R22) and as a negative reference ruderal, dry and anthropogenic vegetation (V38). The functional traits of specific leaf area (SLA; available traits: 99.2% of all species), seed mass (98.8%) and canopy height (99.6%) were retrieved from the databases LEDA [@Kleyer.2008] and TRY [@Kattge.2020].

@@

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists of each plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used 4 chains, a thinning rate of 2, 5000 iterations for warm-up and 10,000 in total. For evaluating the computation, MCMC chain convergence was checked with trace plots and by the R-hat values and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive check was done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019].

A non-metric multidimensional scaling ordination (NMDS) with Jaccard dissimilarity was used to visualize variation in species composition in space and time. Seven species which had an accumulated cover over all plots of less than 0.5% were excluded (2D-stress 0.21).

Data, code and the whole model specification and evaluation is stored on Zenodo and in an easily accessible a scroll-down document on GitHub [!cite Zenodo!]<!--# Cite Zenodo--> There, the sections are referenced to the Bayesian analysis reporting guidelines [BARG, @kruschke2021]. All analyses were performed in R [Version 4.2.2, @base] with the functions 'brm' from the package 'brms' [@brms] for model calculation, several functions from 'brms' and 'bayesplot' [@bayesplot] for model evaluation, 'metaMDS' from 'vegan' for the ordination [@vegan] and 'dbFD' from 'FD' for the calculation of community-weighted means of functional traits [@FD].

@@

We could identify a statistically clear positive effect of sand addition on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). Furthermore, sown hay meadows were clearly closer to their reference plots than sown dry grasslands were to their references. The posterior distributions which are shown in the interactions plots which separates exposition and survey year indicate for all three responses a strong positive development after one year with slow increase in following years (Figures \@ref(fig:persistence); \@ref(fig:fcs); \@ref(fig:recovery)) . Both expositions have similar trends but for all responses the values are clearly lower on the south exposition, e.g., persistence values are on average more than 46% higher on the north exposition (Figure \@ref(fig:persistence)). The effect of restoration treatments is in the interaction plots not clear. Hay meadow-seed mixture led to a closer development to hay meadow references and almost reached on the north exposition the area of the standard deviation of reference sites (Figure \@ref(fig:recovery)). Though, the sand admixture had no clear effect on the development. Persistence of both seed mixture types seems to be slightly positive affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:persistence)).

The ordination visualizes the species composition of seed mixtures and the development of the plots over time (Figure \@ref(fig:nmds)). The NMDS confirms that the seed mixtures are variable, but distinctive for hay meadows and dry grasslands. In addition, it visualizes the decreasing recovery rate on the NMDS1 axis which was already presented (Figure \@ref(fig:recovery)) but the right direction of the development. One exception is that the development of intended hay meadows on the south exposition does not develop towards its seed mixture compositions. The ordination presents the experimental plots not only in context with seed mixtures, but also with reference sites. The reference sites have a bigger variation than the seed mixtures and are close to the seed mixtures but hardly overlap. The positions of reference sites are shifted to the left on the NMDS1 axis which means in the direction of early successional states. On the south exposition, we had negative reference sites which are located in the NMDS in between the positive reference sites and the recent state of restored plots. However, most of negative reference sites (ruderal, dry and anthropogenic vegetation) are situated in the upper half and explicitly recent restoration sites on the south exposition are close to them.

@@

As expected, sand admixture led to higher persistence of sown species but only on the north exposition and the effect size of about 6% in the fourth year after restoration is rather small. In addition, 50% admixture did not further improve the result. Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but can lead to positive results, if mixed with mineral material as sand and gravel [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or Favorable Conservation Status (FCS) similarly to earlier studies [@baer2004]. Larger differences of soil depth might be necessary to create negative effects by greater soil depths as was shown for prairies [@dornbush2010]. Neither, seeding density had an effect on persistence and FCS which fits to result of @Kaulfuss.2022, who found out that a certain amount of seeds are necessary but higher densities do not further improve establishment of target species but have rather a slight negative effect.

@@

Our aim was to identify a ideal seed mixture-substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the establishment and persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions [Figure @ref(fig:traits); @Oberdorfer.1993; @Oberdorfer.1993b]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. Though, these two studies did not work with an experimental set up of different seed mixture-substrate combinations, but analyzed the result of habitat and biotic filtering after five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the sown species could establish earlier than others and preempted the available niches [@Fukami.2015].

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by sowing hay meadows on the north exposition and not on the south exposition of dikes. Furthermore, to reduce nutrient load by sand admixture has a clear positive effect, although the effect is small. The question remains if money is best invested in sand admixture or in other restoration measures. It seems not to be necessary to increase seeding density on dike slopes, which would reduce restoration costs. As well as the use of more deeper soil depth which would otherwise create costs for removal of topsoil. There was no perfect seed mixture-substrate combination, why we conclude that a variation of seed mixtures types and different soils along restoration sections would promote biodiversity better than a single solution [@bauer2023; @Holl.2022]. The strong and lasting negative effect of drought on the south exposition might require a resowing. To close the recovery debt, the adaption of the management might be promising since it is a relevant factor aside of restoration approach and site characteristics [@Grman.2013; @tölgyesi2021]. For example, starting sheep grazing on the restored sites will modify the disturbance regime and improve dispersal. All in all, our results support the finding that dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. Though, the recovery debt highlights that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].

@@

```{r tab.id = "seedmix", tab.cap = "Each plot received an individual set of 20 species with some restrictions to the number of species per functional group. The total species pool for hay meadows was 55 and for dry grassland 58. In total, seed mixtures had 16 or 32 g per plot. All individual seed mixtures are stored in Table A3."}
read_csv(here::here("data", "raw", "data_raw_table_1_seedmix.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  merge_at(part = "header", j = 2:3) %>%
  align(align = "center", part = "all", j = -1) %>%
  add_header_row(top = FALSE, values = c("", "Hay meadow", "Dry grassland", "", "", "")) %>%
  add_header_row(top = FALSE, values = c("", "#", "#", "#", "%", "%")) %>%
  bold(part = "header", i = 1) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  autofit()
```

@@

## Table 2 {.unnumbered}

```{r tab.id = "substrate", ft.align = "left", tab.cap = "Soil of the three substrates was analyzed for the fraction < 2mm fraction. Soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl2 solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in CaCl2 extract. For calculating CaCO3, a subsample was annealed at 550 °C and the measured C amount calculated with 8.33. To calculate total N and the C/N ratio, a subsample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand."}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  select(-starts_with("lfl")) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      select(-starts_with("lfl")) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(i = 1, j = 10, part = "header", value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))) %>%
  compose(i = 1, j = 11, part = "header", value = as_paragraph("K", as_sub("2"), "O")) %>%
  compose(i = 1, j = 12, part = "header", value = as_paragraph("Mg", as_sup("2+"))) %>%
  compose(i = 1, j = 14, part = "header", value = as_paragraph("CaCO", as_sub("3"))) %>%
  compose(i = 2, j = 10, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 11, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 12, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r fig.id = "map", fig.asp = 1.2854, fig.width = 5, fig.cap = "The experiment was located on a dike at river Danube in SE Germany.  The 288 plots were allocated in six blocks (red squares on upper photo) and on the north and south exposition [central photo; both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted (Sand admixture, substrate depth, seeding density, seed mixture types H and D). The western half of a block was with a shallow substrate depth for the blocks and within this half the substrates. The photo on the bottom shows one block from the side after four years (Photo: Markus Bauer)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_1_map.png")
  )
```

>>>>

```{r set-up, include = FALSE, message = FALSE}
library(here)
library(knitr)
library(officedown)
library(officer)
library(tidyverse)
library(flextable)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  dpi = 300,
  fig.cap = FALSE
  )
set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 11,
  text.align = "center"
  )
ft <- officer::fp_text(shading.color = "yellow")

word_spec <- function(x, prop = ft) ftext(text = toString(x) ,prop =  ft)
```

@@

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Though, adapted seed mixtures to certain site conditions promise a more effective restoration. This is especially important if the restored vegetation has to protect constructions like dikes. An ideal seed mixture-substrate combination would concurrently improve the ecosystem function erosion protection and enhance species-rich grasslands by a fast and successful establishment of target species.
2.  An experiment with 288 plots was conducted on both expositions of a dike and was surveyed 2018--2021. We tested different sand admixtures, substrate depths, seed densities and seed mixture types (hay meadows, dry calcareous grasslands), and measured the persistence of sown species, the Favorable Conservation Status (FCS) and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a significant but slightly positive effect on all indices. Substrate depth, seed density and seed mixture type had no significant effect on persistence or FCS.
4.  On the north exposition, hay meadow-seed mixtures developed closer to its reference than dry grasslands. Overall, the vegetation developed in the desired direction, but a recovery debt remained after four years. Plots still exhibited similarities to negative references from ruderal vegetation.
5.  The south exposition was significantly worse than the north exposition which might be due to the severe drought during establishment and the addition and removal of a geotextile. This initial negative effect remained over the whole observation period.
6.  Synthesis and applications: Adapted seed mixtures can make restorations more efficient like applying hay meadows on the north exposition. However, varying seed mixture-substrate combinations might foster biodiversity like beta diversity more than a desired 'ideal' combination. Furthermore, additional efforts in management might be necessary to reduce the recovery debt and resowing after a severe drought might be sensible due to strong legacy effects.

@@

levee

@@

Grasslands provide several ecosystem services and a high biodiversity [@Dengler.2014; @Bardgett.2021]. However, they are globally endangered [@Bardgett.2021] even by restoration through unsuited afforestation [@dudley2020]. In Europe, calcareous grasslands and hay meadows are on the red list [Category 3, vulnerable, @Janssen.2016]. Restoration is seen as a key factor to sustain ecosystem services and biodiversity [@COP.2014; @UnitedNations.2019b] and sowing is a long-established restoration approach for species-rich grasslands [@Kiehl.2010]. Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts [@Freitag.2021]. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021].

@@

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in an agricultural landscape [@Batori.2020]. Dikes have specific characteristics with strongly exposed slopes and river and land orientation [@Batori.2016] and need a vegetation cover to ensure dike stability [@berendse2015]. Dikes can reconcile several requirements [@teixeira2022], which can be fostered by an adapted seed mixture-substrate combination.

@@

The seed mixture-substrate combinations were tested with an experiment on a dike of the River Danube in SE Germany (Figure \@ref(fig:map); 314 m a.s.l.; WGS84 (lat/lon), 48.83895, 12.88412). The climate is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], and three minor floods which did not reach the plots [Appendix A1, @lfu2021waterlevel; @lfu2021discharge]. For the substrates, sand (0--4 mm) and agricultural soil were used and the soil was from a dike construction site nearby (Steinkirchen). A big roller mixed both components and an excavator put the substrates in the prepared plots.

The target vegetation were lowland hay meadows and semi-dry calcareous grassland [EUNIS code: R22, R1A, @Chytry.2020; Arrhenatherion elatioris, Cirsio-Brachypodion pinnati; EuroVegChecklist: CM01A, DA01B, @mucina2016]. The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds. From these species pools, 20 species were randomly selected for each plot (Appendix A2). The seed mixtures always contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Figure \@ref(fig:traits)). As expected, the hay-meadow mixtures had a higher specific leaf area (SLA) lower seed mass and higher canopy height than the dry-meadow mixtures. The south exposed plots were sown in mid-April 2018 and the north exposition 14 days later. Due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks. The management started with a cut without hay removal and a mowing height of 20 cm in August of the same year (2018), followed by standard deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions.

We used 288 plots of the size 2.0 m × 2.0 m which were halfway up the slopes of the dike, distributed over the north and south exposition, and arranged in six blocks (= replicates). The experiment used a split-plot design combined with a randomized complete block design (Figure \@ref(fig:map)). The split plot was created by the two expositions of the dike, where all 24 treatment combinations were tested: sand admixtures (0%, 25% and 50%), soil depths (15 vs. 30 cm) of practical interest [@KleberLerchbaumer.2017], and the two seed mixture types with the normal seed density and an increased density recommended for slopes [4 vs. 8 g m^-2^ @Kiehl.2010; @KleberLerchbaumer.2017].

Under the substrate, a 5 cm thick drainage layer of gravel (0-16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced Calciumcarbonat, but did hardly changed the quite high pH values (Table \@ref(tab:substrate)). Phosphate and potassium were rather scarce for agricultural soils and Magnesium rather high [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A3. `r word_spec("(Add mean values, range, how many of species pool established, etc.)")` <!--# add more infos -->

@@

Reference vegetation surveys were extracted from sPlotOpen [@Sabatini.2021] and our own surveys on Danube dikes in the surroundings [@bauer2023]. We selected six calcareous grassland plots [EUNIS code R1A, @Chytry.2020] within SE Germany from sPlotOpen and 98 plots of our own survey which included also lowland hay meadows (R22) and as a negative reference ruderal, dry and anthropogenic vegetation (V38). The functional traits of specific leaf area (SLA), seed mass and canopy height (available traits of all species: 99.2%, 98.8%, 99.6%) were retrieved from the databases LEDA [@Kleyer.2008] and TRY [@Kattge.2020].

@@

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists of each plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used 4 chains, a thinning rate of 2, 5000 iterations for warm-up and 10,000 in total. For evaluating the computation, MCMC chain convergence was checked with trace plots and by the R-hat values, and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive checks were done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019]. A non-metric multidimensional scaling ordination (NMDS) with Jaccard dissimilarity was used to visualize variation in species composition in space and time (2D-stress 0.21). Seven species were excluded because they had an accumulated cover over all plots of less than 0.5%.

Data, code and the whole model specification and evaluation is stored on Zenodo and in an easily accessible scroll-down document on GitHub `r word_spec("(include Zenodo reference)")`.<!--# Cite Zenodo--> There, the sections are referenced to the Bayesian analysis reporting guidelines [BARG, @kruschke2021]. All analyses were performed in R [Version 4.2.2, @base] with the functions 'brm' from the package 'brms' [@brms] for model calculation, several functions from 'brms' and 'bayesplot' [@bayesplot] for model evaluation, and 'metaMDS' from 'vegan' for the ordination [@vegan].

@@

## Weak effects of restoration treatments

We could identify a statistically clear positive effect of sand admixture on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). The posterior distributions are also shown in the interaction plots which separate exposition and survey year indicate. For all three response variables, the vegetation developed strongly positive after one year, but the recovery rate slowed down in the following years (Figures \@ref(fig:persistence), \@ref(fig:fcs), \@ref(fig:recovery)). Both expositions have similar trends but for all responses the values are clearly lower on the south exposition, e.g., persistence values are on average more than 46% higher on the north exposition (Figure \@ref(fig:persistence)). In the interaction plots, the effects of restoration treatments are neither clear nor strong (Figures \@ref(fig:persistence), \@ref(fig:fcs), \@ref(fig:recovery)). Persistence of both seed mixture types seems to be slightly positive affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:persistence)).

## Hay meadows on north exposition closer to reference

The ordination visualizes the species composition of seed mixtures and the development of the plots over time (Figure \@ref(fig:nmds)). The NMDS confirms that the seed mixtures are variable, but distinctive for hay meadows and dry grasslands and ascertain the right direction of the vegetation development. One exception is that the development of intended hay meadows on the south exposition does not develop towards its seed-mixture compositions.

The ordination presents the experimental plots also in context with reference sites. The reference sites have a bigger variation than the seed mixtures and are close to the seed mixtures but hardly overlap. The positions of reference sites are shifted to the left in comparison to the seed mixtures which means in the direction of early successional states. Though, they are still different from the negative references of ruderal, dry, anthropogenic vegetation. Negative references were only available on the south exposition and they are located in the NMDS between the positive reference sites and the state of restored plots in 2021. Hay meadow-seed mixtures led to a closer development to hay meadow references than dry grasslands to their references (Figure \@ref(fig:effects), \@ref(fig:recovery)) this is especially the case on the north exposition (Figure \@ref(fig:nmds)).

@@

As expected, sand admixture led to higher persistence of sown species but only on the north exposition and the effect size of about 6% in the fourth year after restoration is rather small. In addition, 50% admixture did not further improve the result. Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but can lead to positive results, if mixed with mineral material as sand and gravel [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or Favorable Conservation Status (FCS) similarly to earlier studies [@baer2004]. Larger differences of soil depth might be necessary to create negative effects by greater soil depths as was shown for prairies [@dornbush2010]. Neither, seed density had an effect on persistence and FCS which fits to result of @Kaulfuss.2022, who found out that a certain amount of seeds are necessary but higher densities do not further improve establishment of target species but have rather a slight negative effect.

@@

Our aim was to identify a ideal seed mixture-substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the establishment and persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions [Figure @ref(fig:traits); @Oberdorfer.1993; @Oberdorfer.1993b]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these two studies did not work with an experimental set up of different seed mixture-substrate combinations, but analyzed the result of habitat and biotic filtering after five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the sown species could establish earlier than others and preempted the available niches [@Fukami.2015].

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by sowing hay meadows on the north exposition and not on the south exposition of dikes. Furthermore, to reduce nutrient load by sand admixture has a clear positive effect, although the effect is small. The question remains if money is best invested in sand admixture or in other restoration measures. It seems not to be necessary to increase seed density on dike slopes, which would reduce restoration costs. As well as the use of more deeper soil depth which would otherwise create costs for removal of topsoil. There was no perfect seed mixture-substrate combination, why we conclude that a variation of seed mixtures types and different soils along restoration sections would promote biodiversity better than a single solution [@bauer2023; @Holl.2022]. The strong and lasting negative effect of drought on the south exposition might require a resowing. To close the recovery debt, the adaption of the management might be promising since it is a relevant factor aside of restoration approach and site characteristics [@Grman.2013; @tölgyesi2021]. For example, starting sheep grazing on the restored sites will modify the disturbance regime and improve dispersal. All in all, our results support the finding that dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. Though, the recovery debt highlights that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].

@@

```{r tab.id = "seedmix", ft.align = "left", tab.cap = "Each plot received an individual set of 20 species with some restrictions to the number of species per functional group. The total species pool for hay meadows was 55 and for dry grassland 58. In total, plots received 16 or 32 g of seeds. All individual seed mixtures are stored in Appendix A3."}
read_csv(here::here("data", "raw", "data_raw_table_1_seedmix.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  merge_at(part = "header", j = 2:3) %>%
  align(align = "center", part = "all", j = -1) %>%
  add_header_row(top = FALSE, values = c("", "Hay meadow", "Dry grassland", "", "", "")) %>%
  add_header_row(top = FALSE, values = c("", "#", "#", "#", "wt%", "wt%")) %>%
  bold(part = "header", i = 1) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  autofit()
```

@@

## Table 2<!--# Check quotation --> {.unnumbered}

```{r tab.id = "substrate", ft.align = "left", tab.cap = "Soil of the three substrates was analyzed for the fraction < 2 mm. Soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl2 solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in CaCl2 extract. For calculating CaCO3, a subsample was annealed at 550 °C and the measured C amount calculated with 8.33. To calculate total N and the C/N ratio, a subsample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand."}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  select(-starts_with("lfl")) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      select(-starts_with("lfl")) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(i = 1, j = 10, part = "header", value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))) %>%
  compose(i = 1, j = 11, part = "header", value = as_paragraph("K", as_sub("2"), "O")) %>%
  compose(i = 1, j = 12, part = "header", value = as_paragraph("Mg", as_sup("2+"))) %>%
  compose(i = 1, j = 14, part = "header", value = as_paragraph("CaCO", as_sub("3"))) %>%
  compose(i = 2, j = 10, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 11, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 12, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r fig.id = "map", fig.asp = 1.2854, fig.width = 5, fig.cap = "The experiment was located on a dike at river Danube in SE Germany.  The 288 plots were allocated in six blocks (red squares on upper photo) and on the north and south exposition [central photo; both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted (Sand admixture, substrate depth, seed density, seed mixture types H and D). The western half of a block was with a shallow substrate depth for the blocks and within this half the substrates. The photo on the bottom shows one block from the side after four years (Photo: Markus Bauer)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_1_map.png")
  )
```
